\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage{cvpr}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage[table]{xcolor}
\usepackage{float}
\usepackage{placeins}
\usepackage{caption}

% Include other packages here, before hyperref.

% If you comment hyperref and then uncomment it, you should delete
% egpaper.aux before re-running latex.  (Or just hit 'q' on the first latex
% run, let it finish, and you should be clear).
\usepackage[breaklinks=true,bookmarks=false]{hyperref}

\cvprfinalcopy % *** Uncomment this line for the final submission

\def\cvprPaperID{****} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

% Pages are numbered in submission mode, and unnumbered in camera-ready
%\ifcvprfinal\pagestyle{empty}\fi
%\setcounter{page}{4321}
\begin{document}

%%%%%%%%% TITLE
\title{Industrial Anomaly Detection with Localization: A Comparative Study under Clean and Shifted Domains}

\author{Ivan Necerini \\ s345147\\
{\tt\small s345147@studenti.polito.it}
\and
Jacopo Rialti \\ s346357\\\
{\tt\small s346357@studenti.polito.it}
\and
Fabio Veroli \\ s336301\\
{\tt\small s336301@studenti.polito.it}
}

\maketitle
%\thispagestyle{empty}

%%%%%%%%% ABSTRACT
\begin{abstract}
Visual anomaly detection in manufacturing requires identifying defective products from images acquired under controlled conditions. While state-of-the-art methods such as PatchCore and PaDiM achieve strong results on standard benchmarks, their robustness to distribution shift and behavior in multi-class settings remain underexplored. In this work, we present a comparative study of these two embedding-based approaches on the MVTec~AD dataset under both clean and synthetically shifted domains. We construct MVTec-Shift by applying photometric and geometric transformations inspired by MVTec~AD~2, measuring performance degradation when models trained on clean data are tested under distribution shift. Our experiments reveal that both methods suffer severe specificity collapse without adaptation, with PaDiM exhibiting greater sensitivity to rotations due to its position-specific Gaussian modeling. We evaluate two adaptation strategies: threshold-only recalibration recovers PatchCore's specificity at zero computational cost, while full model retraining provides stronger recovery. Additionally, our coreset ratio ablation shows that aggressive subsampling (1\%) outperforms larger ratios under domain shift, acting as an implicit noise filter. Finally, in the Model-Unified setting, PatchCore maintains robust performance with negligible degradation, while PaDiM exhibits cross-class interference characteristic of the identical shortcut problem. These findings provide practical insights for deploying anomaly detection systems in real-world industrial environments.
\end{abstract}

%%%%%%%%% BODY TEXT
% Include all sections from separate files
\input{sections/01_introduction}
\input{sections/02_related_work}
\input{sections/03_methodology}
\input{sections/04_experimental_setup}
\input{sections/05_results_and_discussion}
\input{sections/06_conclusions.tex}

{\small
\bibliographystyle{ieee_fullname}
\bibliography{egbib}
}

\end{document}
